#### MATH42815 - Machine Learning - 2023/24

#### MATH42815 - Machine Learning - 2023/24

## Machine Learning - Workshop 12


This workbook contains a multi-class classification example with sample code for CART, Random Forests, and Neural Nets.

#### Multi-class classification

The [iris](https://en.wikipedia.org/wiki/Iris_flower_data_set) dataset contains measurements for 150 samples of irisis. There are three species (50 each) in the dataset and 4 measurements for each (Sepal Length in cm, Sepal Width in cm, Petal Length in cm, Petal Width in cm).


```{r}
library(ggplot2)
library(GGally)
data(iris)
head(iris)
iris_results = iris #dataframe to add columns with model results
```


```{r}
ggpairs(iris, aes(color = Species)) + theme_bw()
```


```{r}
library(ggparty)
library(rpart)
```


```{r}
part_species = rpart(data = iris,
                     Species ~ .,
                     method = "class")
part_species
```


```{r}
plot(as.party(part_species))
```


```{r}
printcp(part_species)
autoplot(as.party(part_species))
```


```{r}
check_fit = predict(part_species, newdata = iris, type="class") #change type to prob to see probabilities
#check_fit
table(check_fit, iris$Species) #almost perfect, now split the data in training and testing before fitting and see if it still works
```


```{r}
iris_results$cart = check_fit
head(iris_results)
```


```{r}
library(randomForest)
rf_iris = randomForest(Species ~ ., data = iris)
rf_iris
```


```{r}
iris_results$rf = predict(rf_iris, newdata = iris)
table(iris_results$rf, iris_results$Species)
```


```{r}
library(torch)
train = torch_tensor(as.matrix(iris[,-5]), dtype = torch_float())
target = torch_tensor(as.numeric(iris[,5]), dtype = torch_long())

model_iris = nn_sequential(

  # Layer 1
  nn_linear(4, 8),
  nn_relu(), 

  # Layer 2
  nn_linear(8, 16),
  nn_relu(),

  # Layee 3
  nn_linear(16,3),
  nn_softmax(2)
)
```


```{r}
criterion = nn_cross_entropy_loss()  
optimizer = optim_adam(model_iris$parameters, lr = 0.01)

epochs = 200

# Train the net
for(i in 1:epochs){

  optimizer$zero_grad()

  y_pred = model_iris(train)
  loss = criterion(y_pred, target)
  loss$backward()
  optimizer$step()


  # Check Training
  if(i %% 50 == 0){

    winners = y_pred$argmax(dim=2)
    corrects = (winners == target)
    accuracy = corrects$sum()$item() / target$size()

    cat(" Epoch:", i,"Loss: ", loss$item()," Accuracy:",accuracy,"\n")
  }

}
```


```{r}
#levels(iris$Species)
iris_results$nn = cut(as.numeric(model_iris(train)$argmax(dim=2)), 3, labels = levels(iris$Species))
table(iris_results$nn, iris_results$Species)
```


```{r}
iris_results$cart_right = (iris_results$cart == iris_results$Species)
iris_results$rf_right = (iris_results$rf == iris_results$Species)
iris_results$nn_right = (iris_results$nn == iris_results$Species)
head(iris_results)
```


```{r}
ggplot(iris_results, aes(x=Sepal.Length, y=Sepal.Width))+
geom_point(aes(color = Species,shape = cart_right),size = 2)+
scale_shape_manual(values=c(4, 16)) +
theme_bw()
ggplot(iris_results, aes(x=Sepal.Length, y=Sepal.Width))+
geom_point(aes(color = Species,shape = rf_right),size = 2)+
scale_shape_manual(values=c(16)) +
theme_bw()
ggplot(iris_results, aes(x=Sepal.Length, y=Sepal.Width))+
geom_point(aes(color = Species,shape = nn_right),size = 2)+
scale_shape_manual(values=c(4, 16))+
theme_bw()
```


```{r}
library(lhs)
library(scales)
random_iris = as.data.frame(randomLHS(1000, 4)) #creates a random sample in a hypercube with 4 variables
names(random_iris) = names(iris)[1:4]
random_iris$Sepal.Length = rescale(random_iris$Sepal.Length, to = range(iris$Sepal.Length))
random_iris$Sepal.Width = rescale(random_iris$Sepal.Width, to = range(iris$Sepal.Width))
random_iris$Petal.Length = rescale(random_iris$Petal.Length, to = range(iris$Petal.Length))
random_iris$Petal.Width = rescale(random_iris$Petal.Width, to = range(iris$Petal.Width))
head(random_iris)
```


```{r}
random_iris$cart = predict(part_species, newdata = random_iris, type="class")
random_iris$rf = predict(rf_iris, newdata = random_iris)
tensor_iris = torch_tensor(as.matrix(random_iris[,1:4]), dtype=torch_float())
random_iris$nn = cut(as.numeric(model_iris(tensor_iris)$argmax(dim=2)), 3, labels = levels(iris$Species))
head(random_iris)
```


```{r}
ggpairs(random_iris, aes(color = cart), columns = c(1:4))
```


```{r}
ggpairs(random_iris, aes(color = rf), columns = c(1:4))
```


```{r}
ggpairs(random_iris, aes(color = nn), columns = c(1:4))
```


```{r}

```
