Before you turn this problem in, make sure everything runs as expected. First, **restart the kernel** (in the menubar, select Kernel$\rightarrow$Restart) and then **run all cells** (in the menubar, select Cell$\rightarrow$Run All).

Make sure you fill in any place that says `YOUR CODE HERE` or "YOUR ANSWER HERE", as well as your name and collaborators below:

```{r}
NAME = "Zehao Qian"
COLLABORATORS = "Zehao Qian"
```

------------------------------------------------------------------------

## Machine Learning - Workshop 1

### Preliminaries

**Aim**: The aim of this workshop is to familiarize yourself with some basic concepts and data science tools in R. You have already used R in Intro to Stats and Python in Intro to Maths and Intro to Computer Science. Here we will introduce you to other tools in R that are frequently used in data science.

**R software**: Here you will work with R via the Durham University cluster in Computer Science. If you want to use R in your own machine, we recommend Rstudio (you probably already installed it during Intro to Stats) or a Jupyter Notebook with a R kernel via Anaconda. You can also use other services such as Google Colab (<https://colab.research.google.com/>), GitHub Codespaces (<https://github.com/features/codespaces>).

### Working with Data

Datasets in R are often stored in special types of lists called dataframes (similar to tables in Excel and Pandas in Python). Dataframes are matrix-like blocks of values where each column represents a variable, and each row represents a case. Unlike generic lists, each sublist must contain the same number of elements.

Say we want to create a table with the top 5 grossing films in the UK since 1989 (not corrected for inflation) based on the data in [Wikipedia](https://en.wikipedia.org/wiki/List_of_highest-grossing_films_in_the_United_Kingdom).

1.  The column headers for our initial table will be **Rank** and **Title**.
2.  The values in **Rank** are the numbers 1 to 5.
3.  The values in **Title** are Star Wars:The Force Awakens, Skyfall, No Time to Die, Spider-Man: No Way Home, and Avatar.

```{r}
movies = data.frame(
  Rank = 1:5,
  Title = c("Star Wars: The Force Awakens",
            "Skyfall",
            "No Time to Die",
            "Spiderman: No Way Home",
            "Avatar")
)
View(movies)
```

Take a look at the [R Data Frames page](https://www.w3schools.com/r/r_data_frames.asp) in W3 Schools for more examples.

In the snippet above, we use the View function instead of print. If using RStudio, View opens the dataset in a new tab that let's you preview your dataset, and filter and sort your data more easily. See [View function in R](https://www.statology.org/view-function-in-r/).

Let's add two more columns to the dataframe to include the **Gross** in millions of pounds to the column after **Title**, and then the **Year** of release between the Rank and Title.

```{r}
movies['Gross']=c(123.3, 102.8, 98.0, 97.2, 96.7)
View(movies)
```

```{r}
movies['Year']=c(2015,2012,2021,2021,2009)
movies=movies[,c('Rank','Title','Year','Gross')]
View(movies)
movies
```

There are many ways of adding new rows to a dataframe. If you only need to add one row to the end, you can do as below:

```{r}
movies[6,]=c(6, "Barbie", 2023, 95.6)
View(movies)
```

If you need to add multiple rows, you should create another dataset with the same columns as the original and **rbind** it to the previous one. Later on we will introduce you to the **tidyverse** which will give you more possibilities to execute tasks like this.

```{r}
movies_newrows = data.frame(
  Rank = 7:8,
  Title = c("Spectre","Avengers:Endgame"),
  Year = c(2015, 2019),
  Gross = c(95.2, 88.7)
)
movies = rbind(movies, movies_newrows)
movies
```

Now let's sort this dataset by **Year** in descending order. To do so without introducing new packages, we use the **order** function, not the sort function! The sort function returns the entries in sorted order but it doesn't work well with dataframes; the order function returns the indices of the sorted entries.

By default sorting is ascending using the **order** function so you need to set the *decreasing* parameter to TRUE.

```{r}
View(
  movies[order(movies$Year,decreasing=TRUE),]
)
```

#### **Exercise 1**

Add the next 4 movies to the dataset and sort it by year in **ascending** order:

-   Top Gun: Maverick - 2022 - 83.7
-   Star Wars: The Last Jedi - 2017 - 82.7
-   Titanic - 1998 - 82.7
-   Avatar: The Way of Water - 2022 - 76.9

```{r}
movies_newrows = data.frame(
  Rank = 9:12,
  Title = c("Top Gun: Maverick","Star Wars: The Last Jedi","Titanic","Avatar: The Way of Water"),
  Year = c(2022, 2017, 1998, 2022),
  Gross = c(83.7, 82.7, 82.7, 76.9)
)
movies = rbind(movies, movies_newrows)

View(
  movies[order(movies$Year),]
)
```

### Dataframes and Matrices

Let's load the **Boston** dataset from the **MASS** library. Check your Intro to Stats workshops as a refresher on how to load libraries and datasets!

The Boston dataset contains information on multiple attributes for suburbs in Boston, Massachusetts [Boston in R Documentation](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html).

```{r}
library(MASS)
data(Boston)
head(Boston, n=5) #The head function shows the first n rows in a dataframe
```

Most of the well-known machine learning models are implemented in R (and Python!), and are relatively easy to use. However, there will be cases where you'll need to prepare and transform your data using techniques you have seen in Intro to Maths for example. There might also be cases where you want to implement your own version of a ML model and modify it; therefore, it is useful to know how to manipulate datasets and matrices in R (and understand how to move between those two data types!).

Let's start with a simple example using linear regression and the Boston dataset.

In Intro to Stats Week 6, you studied linear regression models (check your notes!). You used the **lm** function to fit models. In Intro to Maths, you learned how to manipulate matrices and vectors.

In Intro to Stats, you were trying to fit the model

$Y=X\beta+\epsilon$

and estimate the vector $\beta$.

In Intro to Maths, you learned how to invert matrices and solve systems of linear equations. To estimate $\beta$ using least squares, we want to compute:

$\hat{\beta}=(X^TX)^{-1}X^TY$

where $X^T$ is the transpose of $X$ and $A^{-1}$ represents the inverse of a matrix $A$.

#### **Briefly review Intro to Stats workshops 6 and 7 before continuing**

You can also try your solutions here in this notebook to get familiar with the environment.

#### **Exercise 2** - Fitting a linear regression model with **lm**

Use the **lm** function to fit a linear regression model with *medv* as the response variable and the following variables as the predictors: *rm*, *lstat*, *indus*, and *ptratio*.

Save the estimated values of the regression coefficients to the vector **b_lm**.

```{r}
# Fit the linear regression model
model <- lm(medv ~ rm + lstat + indus + ptratio, data=Boston)

# Save the estimated regression coefficients
b_lm <- coef(model)

# Print b
b_lm
```

Now let's try to calculate the $\beta$ estimates using $\hat{\beta}=(X^TX)^{-1}X^TY$.

To do that, we have to define $X$ and $Y$. $Y$ is the vector containing the values of *medv*. The $X$ matrix is the *design matrix* composed by a unit vector linked to the model intercept ($\beta_0$), and the vectors containing the values of the variables *rm*, *lstat*, *indus*, and *ptratio*.

Let's create Y and X:

```{r}
Y = Boston$medv
Xpart = Boston[,c('rm','lstat','indus','ptratio')]
X = cbind(1,Xpart) #this looks like cheating but it works for dataframes,
#you can also use rep to first create a unit vector!
head(X)
```

Now let's convert $Y$ and $X$ to matrices using the *as.matrix* function.

```{r}
Y = as.matrix(Y)
X = as.matrix(X)
```

To transpose the matrix $X$, we use the **t** function. To invert it, we use the **solve** function. Or if you really want to practice your coding skills, you can implement your own inversion function later on!

To multiply matrices and vectors in R, we use %\*%.

```{r}
trX = t(X)
trXX = trX%*%X
trXX
```

```{r}
inv_trXX = solve(trXX)
inv_trXX
```

#### **Exercise 3**

Continue the calculations above and return a vector of length 5 **b_calc**. Compare your vector to **b_lm**.

```{r}
b_calc = inv_trXX %*% (trX %*% Y)
print(b_calc)
print(b_lm)
```

#### **Exercise 4**

Write down and execute the steps you would take to calculate the coefficients for

*lm(medv\~poly(lstat,2, raw=TRUE), data=Boston)*

using matrices as above.

What does the *raw=TRUE* option do in the *poly* function? Increase the power of the polynomial and check what happens to your matrices.

Note: Remember, you can treat powers of *lstat* as separate feature variables

```{r}
Xpart = poly(Boston$lstat, 2, raw = TRUE)
X = cbind(1, Xpart)

Y = as.matrix(Y)
X = as.matrix(X)

trX = t(X)
trXX = trX %*% X
inv_trXX = solve(trXX)

b_calc = inv_trXX %*% (trX %*% Y)

print(b_calc)

lm_model_poly = lm(medv ~ poly(lstat, 2, raw = TRUE), data = Boston)

print(coef(lm_model_poly))
```
