### **Assignment 1 Notebook**

This is the supporting Jupyter Notebook for Assignment 1. Here we will use the King County (USA) *House Data* dataset available via [Kaggle](https://www.kaggle.com/datasets/shivachandel/kc-house-data) .

The original dataset contains 21 variables and 21613 observations. We will use 11 of those variables and our aim is to model house prices.

```{r}
# ---------------------------------------------------------------------
# clear the environment var area
rm(list = ls())
# clear all plots
graphics.off()
# clear the console area
cat("\014")
# ---------------------------------------------------------------------
# current_directory
current_directory = getwd()
# read_csv
# joint file path
file = file.path(current_directory, "kc_house_data.csv")

HouseData = read.csv(file, header=TRUE)
```

```{r}
HouseData =  subset(HouseData, select = c(id, price, bedrooms, bathrooms,
                    sqft_living, sqft_lot, floors,
                    yr_built, zipcode, lat, long))
dim(HouseData) #this should return 21613, 11
names(HouseData) #this should return id, price, bedrooms, bathrooms, sqft_living, sqft_lot, floors, yr_built, zipcode, lat, lon
```

```{R}
head(HouseData)
```

Using an appropriate function from the **leaps** library, perform forward subset selection to find a suitable set of variables to estimate house prices. Uncomment the second line in this block and replace **NULL** with the function call you have written.

```{r}
library(leaps)
best_forward = regsubsets(price ~ ., data = HouseData, nbest = 1, method = "forward", nvmax = 10)
#print the summary for the best_forward output
summary(best_forward)
```

```{r}
#save the summary to forward_summary
forward_summary = summary(best_forward)
#choose the model with the maximum adjusted R2
forward_adjr2 = which.max(forward_summary$adjr2)
#print the coefficients for the model with the highest adj R2
print(coef(best_forward,forward_adjr2))
```

```{r}
forward_cp = which.min(forward_summary$cp)
print(coef(best_forward,forward_cp))
forward_bic = which.min(forward_summary$bic)
print(coef(best_forward,forward_bic))
```

```{r}
par(mfrow = c(1,3)) # allows for 3 plots to be plotted side by side

plot(forward_summary$cp,
     xlab = "# Predictors", #x-axis label
     ylab = "Cp", #y-axis label
     type = "l", #line plot
     lwd = 2) #line thickness

cp_min = which.min(forward_summary$cp)

#overlay the minimum of cp on the previous plot using the points function
points(cp_min,
       forward_summary$cp[cp_min],
       pch = 4, #cross symbol used
       col = 2, #red colour
       cex = 4, #make it bigger!
       lwd = 2) #make the cross lines thicker

# ADD CODE HERE TO PLOT BIC AND ADJUSTED R^2

plot(forward_summary$bic,
     xlab = "# Predictors", #x-axis label
     ylab = "BIC", #y-axis label
     type = "l", #line plot
     lwd = 2) #line thickness


bic_min = which.min(forward_summary$bic)
points(bic_min,
       forward_summary$bic[bic_min],
       pch = 4, #cross symbol used
       col = 2, #red colour
       cex = 4, #make it bigger!
       lwd = 2) #make the cross lines thicker


plot(forward_summary$adjr2,
     xlab = "# Predictors", #x-axis label
     ylab = "AdjR2", #y-axis label
     type = "l", #line plot
     lwd = 2) #line thickness

adjr2_max = which.max(forward_summary$adjr2)
points(adjr2_max,
       forward_summary$adjr2[adjr2_max],
       pch = 4, #cross symbol used
       col = 2, #red colour
       cex = 4, #make it bigger!
       lwd = 2) #make the cross lines thicker
```

```{R}

```
