### **Assignment 1 Notebook**

This is the supporting Jupyter Notebook for Assignment 1. Here we will use the King County (USA) *House Data* dataset available via [Kaggle](https://www.kaggle.com/datasets/shivachandel/kc-house-data) .

The original dataset contains 21 variables and 21613 observations. We will use 11 of those variables and our aim is to model house prices.

```{r}
# ---------------------------------------------------------------------
# clear the environment var area
rm(list = ls())
# clear all plots
graphics.off()
# clear the console area
cat("\014")
# ---------------------------------------------------------------------
# current_directory
current_directory = getwd()
# read_csv
# joint file path
file = file.path(current_directory, "kc_house_data.csv")

HouseData = read.csv(file, header=TRUE)
```

```{r}
HouseData =  subset(HouseData, select = c(id, price, bedrooms, bathrooms,
                    sqft_living, sqft_lot, floors,
                    yr_built, zipcode, lat, long))
dim(HouseData) #this should return 21613, 11
names(HouseData) #this should return id, price, bedrooms, bathrooms, sqft_living, sqft_lot, floors, yr_built, zipcode, lat, lon
```

```{R}
head(HouseData)
```

Using an appropriate function from the **leaps** library, perform forward subset selection to find a suitable set of variables to estimate house prices. Uncomment the second line in this block and replace **NULL** with the function call you have written.

```{r}
library(leaps)
best_forward = regsubsets(price ~ ., data = HouseData, nbest = 1, method = "forward", nvmax = 10)
#print the summary for the best_forward output
summary(best_forward)
```

```{r}
#save the summary to forward_summary
forward_summary = summary(best_forward)
#choose the model with the maximum adjusted R2
forward_adjr2 = which.max(forward_summary$adjr2)
#print the coefficients for the model with the highest adj R2
print(coef(best_forward,forward_adjr2))
```

```{r}
forward_cp = which.min(forward_summary$cp)
print(coef(best_forward,forward_cp))
forward_bic = which.min(forward_summary$bic)
print(coef(best_forward,forward_bic))
```

```{r}
par(mfrow = c(1,3)) # allows for 3 plots to be plotted side by side

plot(forward_summary$cp,
     xlab = "# Predictors", #x-axis label
     ylab = "Cp", #y-axis label
     type = "l", #line plot
     lwd = 2) #line thickness

cp_min = which.min(forward_summary$cp)

#overlay the minimum of cp on the previous plot using the points function
points(cp_min,
       forward_summary$cp[cp_min],
       pch = 4, #cross symbol used
       col = 2, #red colour
       cex = 4, #make it bigger!
       lwd = 2) #make the cross lines thicker

# ADD CODE HERE TO PLOT BIC AND ADJUSTED R^2

plot(forward_summary$bic,
     xlab = "# Predictors", #x-axis label
     ylab = "BIC", #y-axis label
     type = "l", #line plot
     lwd = 2) #line thickness


bic_min = which.min(forward_summary$bic)
points(bic_min,
       forward_summary$bic[bic_min],
       pch = 4, #cross symbol used
       col = 2, #red colour
       cex = 4, #make it bigger!
       lwd = 2) #make the cross lines thicker


plot(forward_summary$adjr2,
     xlab = "# Predictors", #x-axis label
     ylab = "AdjR2", #y-axis label
     type = "l", #line plot
     lwd = 2) #line thickness

adjr2_max = which.max(forward_summary$adjr2)
points(adjr2_max,
       forward_summary$adjr2[adjr2_max],
       pch = 4, #cross symbol used
       col = 2, #red colour
       cex = 4, #make it bigger!
       lwd = 2) #make the cross lines thicker
```

```{r}
lambdas = 10 ^ seq(-6, 6, by = 0.1)
cv_fit = cv.glmnet(
  as.matrix(HouseData[, !names(HouseData) %in% c("price")]),
  as.matrix(HouseData$price),
  alpha = 1,
  lambda = lambdas,
  # nfolds = 20,
  thresh=1e-10
)
plot(cv_fit)
```

$$
p(y|\vec x)=N(y;\hat y(\vec x),\sigma ^2)=\frac{1}{\sqrt {2 \pi \sigma}}e^{-\frac{1}{2}\frac{(y^{(i)}-\hat y)^2}{\sigma^2}}
$$

```{r}
# Sample data
X = c(1.890244, 3.194744, 2.645156, 1.773481, 2.872592, 2.247813, 2.624901, 2.813008, 2.296636, 2.183676)

# Function to calculate the sum of log p-values
log_likelihood_sum = function(sample, mu, sigma) {
  z_scores = abs(sample - mu) / sigma
  p_values = 2 * pnorm(z_scores, lower.tail = FALSE)
  log_p_values = log(p_values)
  sum(log_p_values)
}

# Estimate sample mean and standard deviation
mu_estimate = mean(X)
sigma_estimate = sd(X)

# Calculate the log likelihood sum for the estimated mu and sigma
log_likelihood = log_likelihood_sum(X, mu_estimate, sigma_estimate)
print(paste("Estimated mu:", mu_estimate))
print(paste("Estimated sigma:", sigma_estimate))
print(paste("Log likelihood sum:", log_likelihood))

# Create a scatter plot
plot(
  X,
  main = "Scatter Plot of X",
  xlab = "Index",
  ylab = "Value",
  pch = 19
)
```
